Clustering feature points

### 2013-05-30 text/clustering.tex
Added new figures to the report

\begin{algorithm}
\caption{Mirror Match Algorithm}
\label{alg-mm}
\begin{algorithmic}
\Require $images$ : set of images, $t$ : float $\in \mathbb{R}$
\State $matches_{init}\gets \varnothing$
\State $matches_{final}\gets \varnothing$
\State $features\gets \varnothing$
\ForAll{$I_i \in images$} \Comment Gather features of all images
	\State $features\gets features \cup getFeatures(I_i)$
\EndFor
\ForAll{$f_i \in features$} \Comment Get best match for all features
	\State $f_m,f_n \gets getTwoNearestNeighbors(features \ {f_i})$
	\State $ratio \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_m)$
	\If{$ratio < t$}
		\State $matches_{init}\gets \left(f_i, f_m\right)$
	\EndIf
\EndFor
\ForAll{$\left(f_i, f_j \right) \in matches_{init}$} \Comment Filter 
matches
\If{$\left(f_j, f_i \right) \in matches_{init} \wedge getImg(f_i) \neq 
getImg(f_j)$}
		\State $matches_{final} \gets (f_i, f_j)$
		\State $matches_{init} \gets matches_{init} \ (f_j, f_i)$
	\EndIf
\EndFor \\
\Return $matches_{final}$
\end{algorithmic}
\end{algorithm}

@@

This representation reduces the problem of partitioning the keypoints 
into groups reduces the problem to that of graph clustering or community 
structure depending on the context. In the literature there are various 
ways of clustering a graph according to different measures of what 
constitutes an optimal partitioning. Traditionally the most used 
clustering algorithms have been K-means and spectral clustering, but in 
recent years a host of new algorithms have been proposed based on both 
Newman's concept of graph modularity\footnote{Introduced in 
\cite{girvan2002}, discussed in \cite{brandes2007} and used in 
\cite{blondel2008} as well as others} as well as information theoretical 
measures\footnote{See for example \cite{rosvall2008}} and the Potts spin 
model from physics\footnote{Used in \cite{ronhovde2009}} just to mention 
a few approaches. Many of the new algorithms differ from K-means 
clustering and Spectral clustering in that they don't require the number 
of expected to clusters to be specified beforehand\footnote{Among the 
aforementioned methods, this is true for \cite{blondel2008} and 
\cite{rosvall2008}}.  Furthermore, on tests done using randomly 
generated graphs with a known partitioning \cite{blondel2008}, 
\cite{rosvall2008} and \cite{ronhovde2009} perform markedly better than 
spectral clustering and K-means\cite{lancichinetti2009}.

\subsection{Clustering feature points using Louvain clustering}

\begin{algorithm}
\caption{Cluster-Match Algorithm}
\label{alg-simple}
\begin{algorithmic}
\Require $images$ : set of images, $n$ : int $\in \mathbb{N}$
\State $matches\gets \varnothing$
\ForAll{$I_i \in images$} \Comment Gather features of all images
	\State $f_i\gets getFeatures(I_i)$
\EndFor
\State $A\gets getAdjecencyMatrix(f_1, f_2,\; \ldots \;, f_n)$
\State $A_{norm}\gets 1 - normalize(A)$
\Comment Shorter distance is higher weight
\State $A_{pruned}\gets pruneEdges(A_{norm},\alpha)$
\State $P\gets cluster(A_{pruned})$ 
\Comment P is a set of partitions
\ForAll{$p \in P$} \Comment p is a set of feature points
	\State $trait\gets \left\{p \cap f_i \mid p \cap f_i \neq 
\varnothing, f_i \in \left\{f_1, f_2,\; \ldots \;, f_n\right\}\right\}$
	\If{$\left\vert trait \right\vert > 1$}
		\State $matches\gets getPositions(trait)$
		\Comment Position of trait in each image
	\EndIf
\EndFor \\
\Return matches
\end{algorithmic}
\end{algorithm}

Algorithm \ref{alg-simple} shows the basic steps taken to find matches 
in two or more images. This algorithm is a simplified version of the 
proposed method completely agnostic of the geometry of the feature 
points in the picture. To better explain the algorithm here is a step by 
step explication of the inputs and functions mentioned:
\begin{itemize}
\item[]{\emph{images}: The images we are matching}
\item[]{\emph{$n$}: The number of weights per row in the adjecency 
matrix.}
\item[]{\emph{getFeatures}: Returns a set of feature points given an 
image. In practice the BRIEF\cite{calonder2010brief} feature point 
detector and descriptor has worked well, but it could easily be replaced 
with any other feature point type.} 
% TODO: include a comparison of SURF, SIFT, etc
\item[]{\emph{getAdjecencyMatrix}: Based on the distance measure used to 
compare the similarity of the feature points a symmetric adjecency 
matrix is returned.}
\item[]{\emph{pruneEdges}: The result of the Louvain clustering depends 
on the density of the graph. To ensure an optimal number of partitions 
we only keep the $n$ highest weights per row and column in the adjecency 
matrix, making sure it's still symmetric afterwards. In practice setting 
$n$ to around four or five yields good results.}
\item[]{\emph{cluster}: Cluster the graph corresponding to the adjecency 
matrix given using the Louvain algorithm in \cite{blondel2008}.}
\item[]{\emph{getPositions}: Given a trait consisting of features in 
several different images, find the positions of traits in each image and 
return them. This could either be done by taking a mean of the feature 
points in an image, or by returning the position of the highest matching 
feature points within the trait. In practice ... TODO}
\end{itemize}

>>>>

To discuss the implementation of \emph{MM} and \emph{MMC} a few notes on 
notation will be helpful. BLAH BLAH!

\subsection{Mirror Match (\emph{MM})}

The central idea behind \emph{MM} is to match features of $n$ images by 
taking every feature from all $n$ images and matching it against every 
other feature from the same set. We can then discard the correspondences 
that match two points within the same image. In algorithm \ref{alg-mm} 
the actual implementation of \emph{MM} is shown.

\begin{algorithm}
\caption{Mirror Match Algorithm (\emph{MM})}
\label{alg-mm}
\begin{algorithmic}
\Require $images$ : set of images, $threshold$ : float $\in \mathbb{R}$
\State $matches_{init}\gets \varnothing$
\State $matches_{final}\gets \varnothing$
\State $features\gets \varnothing$
\ForAll{$I_i \in images$} \Comment Gather features of all images
	\State $features\gets features \cup getFeatures(I_i)$
\EndFor
\ForAll{$f_i \in features$} \Comment Get best match for all features
	\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, features ~ 
\backslash ~ \left\{f_i\right\})$
	\State $ratio \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_m)$
	\If{$ratio < threshold$}
		\State $matches_{init}\gets \left(f_i, f_m\right)$
	\EndIf
\EndFor
\ForAll{$\left(f_i, f_j \right) \in matches_{init}$} \Comment Filter 
matches
\If{$\left(f_j, f_i \right) \in matches_{init} \wedge getImg(f_i) \neq 
getImg(f_j)$}
		\State $matches_{final} \gets (f_i, f_j)$
		\State $matches_{init} \gets matches_{init} \ (f_j, f_i)$
	\EndIf
\EndFor \\
\Return $matches_{final}$
\end{algorithmic}
\end{algorithm}

This algorithm passes over three states: An acquisition state where 
feature points are gathered, a matching state where the initial set of 
correspondences are found and a filtering state where correspondences 
within the same image are removed. In the acquisition state the function 
\emph{getFeatures($I_i$)} returns a set of feature points given an 
image.  In practice the feature used for the experiments have been 
SIFT\cite{lowe2004sift}, but there aren't any reason why other features 
such as SURF\cite{bay2006surf} or BRIEF\cite{calonder2010brief} couldn't 
be used just as well. The matching state uses K-Nearest Neighbors to 
return the two closest neighbors in the set of features for any given 
feature $f_i$ and calculates the ratio between the correspondences as 
proposed in \cite{lowe2004sift}.  Any correspondence with a ratio below 
the threshold supplied will be discarded. Finally in the filter the 
function \emph{getImg($f_i$)} returns the parent image of the feature 
$f_i$ and discards both correspondences that aren't symmetric as well as 
those that match two points in the same image.

\begin{figure}
	\makebox[\textwidth][c]{%
		\begin{subfigure}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/mirror_match_off}
			\caption{Baseline Result}
			\label{fig:unique}
		\end{subfigure}%
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)
		\begin{subfigure}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/mirror_match_with_pruned}
			\caption{\emph{MM} with intra image matches}
			\label{fig:within}
		\end{subfigure}%
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad
		  %(or a blank line to force the subfigure onto a new line)
		\begin{subfigure}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/mirror_match}
			\caption{\emph{MM}}
			\label{fig:without}
		\end{subfigure}%
	}%
	\label{fig:compare_mirror}
	\caption{Illustration of matches filtered by \emph{MM}}
\end{figure}

Figure \ref{fig:compare_mirror} shows an example of the performance of 
\emph{MM}. In sub figure \ref{fig:unique} the result of a normal 
matching of two images is shown. The two images appear side by side and 
for each correspondence between them, a line has been drawn from one 
image to the other. The color of the line indicates if the 
correspondence is accurate. Sub figure \ref{fig:within} shows the 
correspondences found by mirror match using the same threshold before 
the matches within the same image has been filtered out. In particular 
the fence by the bottom of the image admits correspondences that can 
conveniently be removed as done in sub figure \ref{fig:without} where 
only the actual correspondences returned by the algorithm are shown.

\subsection{Mirror Match with Clustering (\emph{MMC})}

As opposed to \emph{MM}, \emph{MMC} diverges from traditional 
non-geometric feature matching by clustering the feature points by 
similarity. This process yields partitions of fairly similar feature 
points that we can match using the same approach as \emph{MM}. Before I 
introduce the implementation details I will go over the problem of graph 
clustering and how it relates to feature matching.

\subsubsection{Graph Clustering}

This representation reduces the problem of partitioning the keypoints 
into groups to that of graph clustering or community structure depending 
on the context. In the literature there are various ways of clustering a 
graph according to different measures of what constitutes an optimal 
partitioning. Traditionally the most used clustering algorithms have 
been K-means and spectral clustering, but in recent years a host of new 
algorithms have been proposed based on both Newman's concept of graph 
modularity\footnote{Introduced in \cite{girvan2002}, discussed in 
\cite{brandes2007} and used in \cite{blondel2008} as well as others} as 
well as information theoretical measures\footnote{See for example 
\cite{rosvall2008}} and the Potts spin model from physics\footnote{Used 
in \cite{ronhovde2009}} just to mention a few approaches. Many of the 
new algorithms differ from K-means clustering and Spectral clustering in 
that they don't require the number of expected to clusters to be 
specified beforehand\footnote{Among the aforementioned methods, this is 
true for \cite{blondel2008} and \cite{rosvall2008}}.  Furthermore, on 
tests done using randomly generated graphs with a known partitioning 
\cite{blondel2008}, \cite{rosvall2008} and \cite{ronhovde2009} perform 
markedly better than spectral clustering and 
K-means\cite{lancichinetti2009}.

\subsubsection{Clustering feature points using Louvain clustering}

\begin{algorithm}
\caption{Mirror Match with Clustering Algorithm (\emph{MMC})}
\label{alg-mmc}
\begin{algorithmic}
\Require $images$ : set of images, $threshold \in \mathbb{R}$
\State $matches\gets \varnothing$
\State $features\gets \varnothing$
\ForAll{$I_i \in images$} \Comment Gather features of all images
	\State $f_i\gets getFeatures(I_i)$
	\State $features\gets features \cup f_i$
\EndFor
\State $A\gets getAdjecencyMatrix(f_1, f_2,\; \ldots \;, f_n)$
\State $A_{norm}\gets 1 - normalize(A)$
\Comment Shorter distance is higher weight
\State $A_{pruned}\gets pruneEdges(A_{norm},\alpha)$
\State $Partitions\gets cluster(A_{pruned})$
\ForAll{$p \in Partitions$} \Comment p is a set of feature points
	\State $matches\gets matches \cup getMatches(p, threshold, 
features)$
\EndFor \\
\Return matches
\end{algorithmic}
\end{algorithm}

Unlike the \emph{MM} algorithm the feature points are partitioned before 
the matching stage. While the louvain clustering algorithm doesn't 
require any parameters in itself, it tends towards clustering all 
feature points together in the same partition if the graph is fully 
connected or close. To ensure that the graph is well clustered, the 
adjacency matrix is pruned so only edges above a certain threshold is 
kept. From empirical analysis keeping around 2.5\% of edges seems to 
work well. Figure \ref{fig:graph} shows the result of clustering the 
feature points displayed as a graph where each node correspond to a 
feature point. Here the border of nodes signifies the original image 
while the color of the node is decided by what partition it belongs to.  
Notice that since a limited amount of colors are used disconnected 
subgraphs of the same color can be assumed to belong to separate 
partitions. The sub figure \ref{fig:cropped_graph} shows a detailed 
section of the graph where the interplay between small clusters 
consisting of two or three nodes as well as larger clusters consisting 
of several hundred nodes can be seen.

\begin{figure}
	\makebox[\textwidth][c]{%
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_graph_full}
			\caption{Full Graph}
			\label{fig:full_graph}
		\end{subfigure}%
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_graph_cropped}
			\caption{Cropped Section of Graph}
			\label{fig:cropped_graph}
		\end{subfigure}%
	}%
	\label{fig:graph}
	\caption{The partitioned feature graph. Every color signifies a 
partition while the edge color of each node signifies which image it 
belongs to}
\end{figure}

The clusters 
\begin{figure}
	\makebox[\textwidth][c]{%
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_pitts_source}
			\caption{Source image}
			\label{fig:pitts_source}
		\end{subfigure}%
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_pitts_keypoints}
			\caption{Keypoints Displayed}
			\label{fig:pitts_keypoints}
		\end{subfigure}%
	}%
	\\
	\makebox[\textwidth][c]{
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)

		\begin{subfigure}[t]{1.16\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_partition}
			\caption{A Partition Example}
			\label{fig:pitts_partition}
		\end{subfigure}%
	}%
	\label{fig:compare_mirror}
	\caption{Example of partitioning of keypoints in an image}
\end{figure}

\begin{algorithm}
\caption{Impl. of GetMatches (\emph{from MMC algorithm})}
\label{alg-getmatches}
\begin{algorithmic}
\Require $p$ : set of features, $threshold \in \mathbb{R}$, $features$ : 
Set of all features
\State $matches \gets \varnothing$
\State $edges \gets \left\{getSimilarity(f_i, f_j) \mid getImg(f_i) \neq 
getImg(f_j) \wedge f_i, f_j \in p \right\}$
\If{$\left\vert edges \right\vert = 1$} \Comment If $\exists$ one edge 
between images
	\State $f_i,f_j \gets getInterImageFeatures(edges_{inter}, p)$
	\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, features ~ 
\backslash ~ \left\{f_i\right\})$
	\State $f_s,f_t \gets getTwoNearestNeighbors(f_j, features ~ 
\backslash ~ \left\{f_j\right\})$
	\State $ratio_i \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_j)$
	\State $ratio_j \gets getSimilarity(f_j, f_t) / getSimilarity(f_j, 
f_i)$
	\If{$ratio_i < threshold \wedge ratio_j < threshold$}
		\State $matches \gets matches \cup (f_i, f_j)$
	\EndIf
\ElsIf{$\left\vert edges \right\vert > 1$} \Comment If $\exists$ more 
than one edge between images
	\ForAll{$f_i \in p$}
		\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, p ~ \backslash 
~ \left\{f_i\right\})$
		\State $ratio \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_m)$
		\If{$getImg(f_i) \neq getImg(f_m) \wedge ratio < threshold 
\wedge (f_m, f_i) \not\in matches$}
			\State $matches \gets matches \cup (f_i, f_m)$
		\EndIf
	\EndFor
\EndIf

\Return matches
\end{algorithmic}
\end{algorithm}

Algorithm \ref{alg-mmc} shows the basic steps taken to find matches in 
two or more images.

This algorithm is a simplified version of the proposed method completely 
agnostic of the geometry of the feature points in the picture. To better 
explain the algorithm here is a step by step explication of the inputs 
and functions mentioned:
\begin{itemize}
\item[]{\emph{images}: The images we are matching}
\item[]{\emph{$n$}: The number of weights per row in the adjecency 
matrix.}
\item[]{\emph{getFeatures}: Returns a set of feature points given an 
image. In practice the BRIEF\cite{calonder2010brief} feature point 
detector and descriptor has worked well, but it could easily be replaced 
with any other feature point type.} 
% TODO: include a comparison of SURF, SIFT, etc
\item[]{\emph{getAdjecencyMatrix}: Based on the distance measure used to 
compare the similarity of the feature points a symmetric adjecency 
matrix is returned.}
\item[]{\emph{pruneEdges}: The result of the Louvain clustering depends 
on the density of the graph. To ensure an optimal number of partitions 
we only keep the $n$ highest weights per row and column in the adjecency 
matrix, making sure it's still symmetric afterwards. In practice setting 
$n$ to around four or five yields good results.}
\item[]{\emph{cluster}: Cluster the graph corresponding to the adjecency 
matrix given using the Louvain algorithm in \cite{blondel2008}.}
\item[]{\emph{getPositions}: Given a trait consisting of features in 
several different images, find the positions of traits in each image and 
return them. This could either be done by taking a mean of the feature 
points in an image, or by returning the position of the highest matching 
feature points within the trait. In practice ... TODO}
\end{itemize}

The two algorithms introduced in this report are designed to work with 
feature points such as ,  or similar. While the experimental results 
have all been done with the SIFT descriptor, there aren't any 
theoretical or practical issues with using other types of feature 
points.

@@
